Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2024-02-22 paper/main.tex
saved out updated fig, fix fonts in illustrator, updated fig 6 stats in paper

When we fit a GLMM to estimates of participants' knowledge for each Quiz~1
question based on all other Quiz~1 questions, we observed an outcome consistent
with our null hypothesis: higher estimated knowledge at the embedding
coordinate of a held-out question was associated with a lower likelihood of
answering the question correctly (odds ratio $(OR) = 0.136$, likelihood-ratio
test statistic $(\lambda_{LR}) = 19.749$, 95\%\ $\textnormal{CI} = [14.352,\
26.545],\ p = 0.001$). This outcome suggests that our knowledge estimates do
\textit{not} provide useful information about participants' Quiz~1 performance
when we aggregated across all question content areas. We speculated that this
might either indicate that the knowledge estimates are uninformative in
general, or about Quiz~1 performance in particular. This would be expected, for
example, if participants were guessing about the answers to the Quiz~1
questions (prior to having watched either lecture). When we repeated this
analysis for Quizzes~2 and~3, we found that \textit{higher} estimated knowledge
for a given question predicted a greater likelihood of answering it correctly
(Quiz~2: $OR = 2.905,\ \lambda_{LR} = 17.333,\ 95\%\ \textnormal{CI} =
[14.966,\ 29.309],\ p = 0.002$; Quiz~3: $OR = 3.238,\ \lambda_{LR} = 6.882,\
95\%\ \textnormal{CI} = [6.228,\ 8.184],\ p = 0.017$). Taken together, these
results suggest that our knowledge estimates reliably predict participants'
performance on individual held-out quiz questions, but only after participants
have received at least some training.

We observed a similar pattern of results when used this approach to estimate
participants' knowledge about held-out questions from one lecture using their
performance on other questions from the \textit{same} lecture. Specifically,
for Quiz~1 questions (i.e., prior to watching either), participants' estimated
knowledge for the embedding coordinates of held-out \textit{Four Fundamental
Forces}-related questions estimated using other \textit{Four Fundamental
Forces}-related questions did not reliably predict whether those questions were
answered correctly ($OR = 1.891,\ \lambda_{LR} = 2.293,\ 95\%\ \textnormal{CI}
= [2.091,\ 2.622],\ p = 0.139$). The same was true of knowledge estimates for
held-out \textit{Birth of Stars}-related questions based on other \textit{Birth
of Stars}-related questions ($OR = 0.722,\ \lambda_{LR} = 5.115,\ 95\%\
\textnormal{CI} = [0.094,\ 0.146],\ p = 0.738$). As in our analysis that
included all questions, we speculate that these ``null'' results might reflect
some degree of random guessing on Quiz~1. When we repeated these within-lecture
analyses using questions from Quiz~2 (which participants took immediately after
viewing \textit{Four Fundamental Forces} but prior to viewing \textit{Birth of
Stars}), we found that they now reliably predicted success on \textit{Four
Fundamental Forces}-related questions ($OR = 9.023,\ \lambda_{LR} = 18.707,\
95\%\ \textnormal{CI} = [10.877,\ 22.222],\ p = 0.001$) but not on
\textit{Birth of Stars}-related questions ($OR = 0.306,\ \lambda_{LR} = 5.115,\
95\%\ \textnormal{CI} = [4.624,\ 5.655],\ p = 0.055$). Here, we speculate that
participants might have been guessing about the \textit{Birth of Stars} content
(e.g., prior to having watched it), whereas they might have been drawing on
some structured knowledge about the \textit{Four Fundamental Forces} content
(e.g., from having just watched it). When we applied this approach to Quiz~3
responses (given immediately after viewing \textit{Birth of Stars}), we found
that within-lecture knowledge estimates for \textit{Birth of Stars}-related
questions could now reliably predict success on those questions ($OR = 5.467,\
\lambda_{LR} = 10.670,\ 95\%\ \textnormal{CI} = [7.998, 12.532],\ p = 0.006$).
However, within-lecture knowledge estimates for \textit{Four Fundamental
Forces} questions answered on Quiz~3 were no longer directly related to the
likelihood of successfully answering them and instead exhibited the inverse
relationship we would expect to arise from unstructured knowledge (with respect
to the embedding space; $OR = 0.013,\ \lambda_{LR} = 14.648,\ 95\%\
\textnormal{CI} = [10.695, 23.096],\ p = 0.001$). Speculatively, we suggest
that this may reflect participants forgetting some of the \textit{Four
Fundamental Forces} content (e.g., perhaps in favor of prioritizing encoding
the just-watched \textit{Birth of Stars} content in preparation for the third
quiz). If this forgetting happens in a relatively ``random'' way (with respect
to spatial distance within the embedding space), then it could explain why some
held-out questions about \textit{Four Fundamental Forces} were answered
incorrectly, even if questions at nearby coordinates (i.e., about similar
content) were answered correctly. This might lead our approach to over-estimate
knowledge for held-out questions about ``forgotten'' knowledge that
participants answered incorrectly. Taken together, these within-lecture results
suggest that our approach can distinguish between questions about different
content covered by a single lecture when participants have sufficiently
structured knowledge about its contents, though this specificity may decrease
with time since the relevant material was learned.

Finally, we used this approach to estimate participants' knowledge about
held-out questions from one lecture using their performance on questions from
the \textit{other} lecture. Here we again observed a similar pattern of
results, though with some notable differences. On Quiz~1, we found that
participants' abilities to correctly answer questions about \textit{Four
Fundamental Forces} could be predicted from their responses to questions about
\textit{Birth of Stars} ($OR = 1.896,\ \lambda_{LR} = 7.205,\ 95\%\
\textnormal{CI} = [6.224, 7.524],\ p = 0.039$) and similarly, that their
ability to correctly answer \textit{Birth of Stars}-related questions could be
predicted from their responses to \textit{Four Fundamental Forces}-related
questions ($OR = 1.522,\ \lambda_{LR} = 6.448,\ 95\%\ \textnormal{CI} = [5.656,
6.843],\ p = 0.043$). Given the results from our analyses that included all
questions and within-lecture predictions, we were surprised to find that the
knowledge estimates could reliably (if weakly) predict participants'
performance across content from different lectures. It is possible that this
result reflects a combination of random guessing prior to training (leading to
a weak effect size), alongside some coarse-scale structured knowledge that
participants had about the content prior to watching either lecture. When we
repeated this analysis using questions from Quiz~2, we found participants'
responses to \textit{Four Fundamental Forces}-related questions did
\textit{not} reliably predict their success on \textit{Birth of Stars}-related
questions ($OR = 1.865,\ \lambda_{LR} = 3.205,\ 95\%\ \textnormal{CI} = [3.027,
3.600],\ p = 0.125$), nor did their responses to \textit{Birth of
Stars}-related questions reliably predict their success on \textit{Four
Fundamental Forces}-related questions ($OR = 3.490,\ \lambda_{LR} = 3.266,\
95\%\ \textnormal{CI} = [3.033, 3.866],\ p = 0.094$). These ``prediction
failures'' appear to come from the fact that any signal derived from
participants' knowledge about the content of the \textit{Birth of Stars}
lecture (prior to watching it) is swamped by the much more dramatic increase in
their knowledge about the content of the \textit{Four Fundamental Forces}
(which they watched just prior to taking Quiz~2). This is reflected in their
Quiz~2 performance for questions about each lecture (mean proportion correct
for \textit{Four Fundamental Forces}-related questions on Quiz~2: $0.77$; mean
proportion correct for \textit{Birth of Stars}-related questions on Quiz~2:
$0.36$). When we carried out these across-lecture knowledge predictions using
questions from Quiz~3 (when participants had now viewed \textit{both}
lectures), we could again reliably predict success on questions about both
\textit{Four Fundamental Forces} ($OR = 11.294),\ \lambda_{LR} = 11.055,\ 95\%\
\textnormal{CI} = [9.126, 18.476],\ p = 0.004$) and \textit{Birth of Stars}
($OR = 7.302,\ \lambda_{LR} = 7.068,\ 95\%\ \textnormal{CI} = [6.490, 8.584],\
p = 0.017$) using responses to questions about the other lecture's content.
Across all three versions of these analyses, our results suggest that (by and
large) our knowledge estimates can reliably predict participants' abilities to
answer individual quiz questions, distinguish between questions about similar
content, and generalize across content areas, provided that participants' quiz
responses reflect a minimum level of ``real'' knowledge about both content on
which these predictions are based and that for which they are made. Our results
also indicate some important limitations of our approach: if participants' quiz
performance does not reflect what they know (e.g., when they ``guess''), or if
their knowledge is not structured in a way that is reflected by the embedding
space, then our knowledge estimates will not be predictive of their
performance.

% our approach works when participants have a minimal baseline level of knowledge about content predicted and used to predict
% our approach generalizes when knowledge of content used to predict can be assumed to be a reasonable indicator of knowledge of content predicted
% our approach has enough specificity to distinguish between content within the same lecture when it was just watched -- maybe when people forget a little bit they forget "randomly"?.

% potential new transition/motivation -- in the previous analyses, we identified a particular set of constraints on our estimates of participants' knowledge. This made us wonder about another potential constraint: how far away in topic space does the relevance of being able to answer a question extend and influence ability to answer a different question?

>>>>

When we fit a GLMM to estimates of participants' knowledge for each Quiz~1
question based on all other Quiz~1 questions, we observed an outcome consistent
with our null hypothesis: higher estimated knowledge at the embedding
coordinate of a held-out question was associated with a lower likelihood of
answering the question correctly (odds ratio $(OR) = 0.136$, likelihood-ratio
test statistic $(\lambda_{LR}) = 19.749$, 95\%\ $\textnormal{CI} = [14.352,\
26.545],\ p < 0.001$). This outcome suggests that our knowledge estimates do
\textit{not} provide useful information about participants' Quiz~1 performance
when we aggregated across all question content areas. We speculated that this
might either indicate that the knowledge estimates are uninformative in
general, or about Quiz~1 performance in particular. This would be expected, for
example, if participants were guessing about the answers to the Quiz~1
questions (prior to having watched either lecture). When we repeated this
analysis for Quizzes~2 and~3, we found that \textit{higher} estimated knowledge
for a given question predicted a greater likelihood of answering it correctly
(Quiz~2: $OR = 2.905,\ \lambda_{LR} = 17.333,\ 95\%\ \textnormal{CI} =
[14.966,\ 29.309],\ p = 0.001$; Quiz~3: $OR = 3.238,\ \lambda_{LR} = 6.882,\
95\%\ \textnormal{CI} = [6.228,\ 8.184],\ p = 0.016$). Taken together, these
results suggest that our knowledge estimates reliably predict participants'
performance on individual held-out quiz questions, but only after participants
have received at least some training.

We observed a similar pattern of results when used this approach to estimate
participants' knowledge about held-out questions from one lecture using their
performance on other questions from the \textit{same} lecture. Specifically,
for Quiz~1 questions (i.e., prior to watching either), participants' estimated
knowledge for the embedding coordinates of held-out \textit{Four Fundamental
Forces}-related questions estimated using other \textit{Four Fundamental
Forces}-related questions did not reliably predict whether those questions were
answered correctly ($OR = 1.891,\ \lambda_{LR} = 2.293,\ 95\%\ \textnormal{CI}
= [2.091,\ 2.622],\ p = 0.138$). The same was true of knowledge estimates for
held-out \textit{Birth of Stars}-related questions based on other \textit{Birth
of Stars}-related questions ($OR = 0.722,\ \lambda_{LR} = 5.115,\ 95\%\
\textnormal{CI} = [0.094,\ 0.146],\ p = 0.738$). As in our analysis that
included all questions, we speculate that these ``null'' results might reflect
some degree of random guessing on Quiz~1. When we repeated these within-lecture
analyses using questions from Quiz~2 (which participants took immediately after
viewing \textit{Four Fundamental Forces} but prior to viewing \textit{Birth of
Stars}), we found that they now reliably predicted success on \textit{Four
Fundamental Forces}-related questions ($OR = 9.023,\ \lambda_{LR} = 18.707,\
95\%\ \textnormal{CI} = [10.877,\ 22.222],\ p < 0.001$) but not on
\textit{Birth of Stars}-related questions ($OR = 0.306,\ \lambda_{LR} = 5.115,\
95\%\ \textnormal{CI} = [4.624,\ 5.655],\ p = 0.054$). Here, we speculate that
participants might have been guessing about the \textit{Birth of Stars} content
(e.g., prior to having watched it), whereas they might have been drawing on
some structured knowledge about the \textit{Four Fundamental Forces} content
(e.g., from having just watched it). When we applied this approach to Quiz~3
responses (given immediately after viewing \textit{Birth of Stars}), we found
that within-lecture knowledge estimates for \textit{Birth of Stars}-related
questions could now reliably predict success on those questions ($OR = 5.467,\
\lambda_{LR} = 10.670,\ 95\%\ \textnormal{CI} = [7.998, 12.532],\ p = 0.005$).
However, within-lecture knowledge estimates for \textit{Four Fundamental
Forces} questions answered on Quiz~3 were no longer directly related to the
likelihood of successfully answering them and instead exhibited the inverse
relationship we would expect to arise from unstructured knowledge (with respect
to the embedding space; $OR = 0.013,\ \lambda_{LR} = 14.648,\ 95\%\
\textnormal{CI} = [10.695, 23.096],\ p < 0.001$). Speculatively, we suggest
that this may reflect participants forgetting some of the \textit{Four
Fundamental Forces} content (e.g., perhaps in favor of prioritizing encoding
the just-watched \textit{Birth of Stars} content in preparation for the third
quiz). If this forgetting happens in a relatively ``random'' way (with respect
to spatial distance within the embedding space), then it could explain why some
held-out questions about \textit{Four Fundamental Forces} were answered
incorrectly, even if questions at nearby coordinates (i.e., about similar
content) were answered correctly. This might lead our approach to over-estimate
knowledge for held-out questions about ``forgotten'' knowledge that
participants answered incorrectly. Taken together, these within-lecture results
suggest that our approach can distinguish between questions about different
content covered by a single lecture when participants have sufficiently
structured knowledge about its contents, though this specificity may decrease
with time since the relevant material was learned.

Finally, we used this approach to estimate participants' knowledge about
held-out questions from one lecture using their performance on questions from
the \textit{other} lecture. Here we again observed a similar pattern of
results, though with some notable differences. On Quiz~1, we found that
participants' abilities to correctly answer questions about \textit{Four
Fundamental Forces} could be predicted from their responses to questions about
\textit{Birth of Stars} ($OR = 1.896,\ \lambda_{LR} = 7.205,\ 95\%\
\textnormal{CI} = [6.224, 7.524],\ p = 0.038$) and similarly, that their
ability to correctly answer \textit{Birth of Stars}-related questions could be
predicted from their responses to \textit{Four Fundamental Forces}-related
questions ($OR = 1.522,\ \lambda_{LR} = 6.448,\ 95\%\ \textnormal{CI} = [5.656,
6.843],\ p = 0.042$). Given the results from our analyses that included all
questions and within-lecture predictions, we were surprised to find that the
knowledge estimates could reliably (if weakly) predict participants'
performance across content from different lectures. It is possible that this
result reflects a combination of random guessing prior to training (leading to
a weak effect size), alongside some coarse-scale structured knowledge that
participants had about the content prior to watching either lecture. When we
repeated this analysis using questions from Quiz~2, we found participants'
responses to \textit{Four Fundamental Forces}-related questions did
\textit{not} reliably predict their success on \textit{Birth of Stars}-related
questions ($OR = 1.865,\ \lambda_{LR} = 3.205,\ 95\%\ \textnormal{CI} = [3.027,
3.600],\ p = 0.124$), nor did their responses to \textit{Birth of
Stars}-related questions reliably predict their success on \textit{Four
Fundamental Forces}-related questions ($OR = 3.490,\ \lambda_{LR} = 3.266,\
95\%\ \textnormal{CI} = [3.033, 3.866],\ p = 0.093$). These ``prediction
failures'' appear to come from the fact that any signal derived from
participants' knowledge about the content of the \textit{Birth of Stars}
lecture (prior to watching it) is swamped by the much more dramatic increase in
their knowledge about the content of the \textit{Four Fundamental Forces}
(which they watched just prior to taking Quiz~2). This is reflected in their
Quiz~2 performance for questions about each lecture (mean proportion correct
for \textit{Four Fundamental Forces}-related questions on Quiz~2: $0.77$; mean
proportion correct for \textit{Birth of Stars}-related questions on Quiz~2:
$0.36$). When we carried out these across-lecture knowledge predictions using
questions from Quiz~3 (when participants had now viewed \textit{both}
lectures), we could again reliably predict success on questions about both
\textit{Four Fundamental Forces} ($OR = 11.294),\ \lambda_{LR} = 11.055,\ 95\%\
\textnormal{CI} = [9.126, 18.476],\ p = 0.003$) and \textit{Birth of Stars}
($OR = 7.302,\ \lambda_{LR} = 7.068,\ 95\%\ \textnormal{CI} = [6.490, 8.584],\
p = 0.016$) using responses to questions about the other lecture's content.
Across all three versions of these analyses, our results suggest that (by and
large) our knowledge estimates can reliably predict participants' abilities to
answer individual quiz questions, distinguish between questions about similar
content, and generalize across content areas, provided that participants' quiz
responses reflect a minimum level of ``real'' knowledge about both content on
which these predictions are based and that for which they are made. Our results
also indicate some important limitations of our approach: if participants' quiz
performance does not reflect what they know (e.g., when they ``guess''), or if
their knowledge is not structured in a way that is reflected by the embedding
space, then our knowledge estimates will not be predictive of their
performance.
